---
title: "Univariate GWAS with wgAim"
author: "Julian Taylor & Ari Verbyla"
date: today
vignette: >
  %\VignetteIndexEntry{Univariate GWAS with wgAim}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 10
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
  cache: true
---

```{r setup}
#| include: false
#| cache: false
asreml_ok <- requireNamespace("asreml", quietly = TRUE)
if (requireNamespace("wgAim", quietly = TRUE)) {
  library(wgAim)
} else {
  devtools::load_all("C:/jtaylor/Research/Computing/R Packages/Claude/wgaim",
                     quiet = TRUE)
}
library(ggplot2)
if (asreml_ok) {
  library(asreml)
  asreml.options(trace = FALSE)
}
```

---

## Overview

This vignette demonstrates the complete univariate **wgAim** GWAS pipeline
for a diversity panel evaluated in a single environment. The pipeline has
three stages:

1. **Pre-analysis** — inspect a raw 0/1/2 marker genotype matrix with
   `checkPanel()`, apply quality filters with `filterPanel()`, and
   assemble the analysis-ready `wgPanel` object with `primePanel()`
   (including KNN imputation of residual missingness).
2. **Analysis** — fit a whole-genome composite marker model and iteratively
   detect associated markers with `gwasAim()`.
3. **Post-analysis** — interrogate, summarise, and visualise the detected
   associations using the `print`, `summary`, and `plot` methods, and
   refine peak positions with `fineMap()`.

Throughout the vignette we use the simulated `genoGWASraw`, `mapGWASsim`
and `phenoGWASsim` datasets bundled with the package. These represent a
diversity panel of 200 lines genotyped at approximately 1700 SNP markers
across 10 chromosomes (about 170 markers per chromosome, average spacing
~0.9 cM). The panel is generated with `AlphaSimR` — 50 founder haplotypes
via a coalescent (`runMacs2`), one generation of random mating to
produce 200 F1 diverse lines, then a SNP chip actively selected from
segregating sites with minor allele frequency (MAF) > 0.15. All shipped
markers have F1 MAF > 0.15 (median ~0.32), so the panel is dominated by
informative common variants. The resulting kinship matrix
$\boldsymbol{K}$ has genuine full-rank structure (off-diagonal SD
~0.085) and captures the half-sib / full-sib relatedness naturally.
The raw marker matrix deliberately contains a range of realistic
data-quality issues (missingness, monomorphic markers, duplicate lines,
duplicate markers and markers absent from the map) so the full
pre-analysis pipeline can be exercised. Five QTL were planted at known
positions to give ground truth for detection. The phenotype is a balanced
6-site, 2-replicate multi-environment trial on the tonnes-per-hectare
scale (site means 4.5–4.9 t/ha, residual SD ~0.377 t/ha per site) with
two Kronecker-structured polygenic components matching the wgAim
multivariate analysis's `fa(Site,k):vm(id, K) + fa(Site,k):id`
decomposition (kept deliberately small so the composite LRT stops
cleanly once true QTL are absorbed); for the single-environment vignette
we use only `Env1`.

::: {.note}
**Requires:** ASReml-R V4. The vignette is pre-rendered so no licence is
needed to read it. Run the code chunks only if you have ASReml-R installed.
:::

---

## Mathematical Framework

The mathematical framework underpinning `gwasAim()` is essentially
identical to that of `qtlAim()`. Please see the **Mathematical
Framework** section of the [Univariate QTL Mapping
vignette](qtlAim_pipeline.html#mathematical-framework) for the full
development, covering:

- the base linear mixed model
  $\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\tau} + \boldsymbol{Z}_u\boldsymbol{u} + \boldsymbol{Z}_g\boldsymbol{g} + \boldsymbol{e}$;
- decomposition of the genetic term $\boldsymbol{g}$ into a whole-genome
  additive component $\boldsymbol{a}$ and a residual polygenic
  background $\boldsymbol{p}$;
- the two computational paths for $\boldsymbol{a}$ — **vm**
  (genomic relationship matrix
  $\boldsymbol{K} = \tfrac{1}{p}\boldsymbol{M}\boldsymbol{M}^{\top}$
  when $p > n$) and **mbf** (ridge regression on marker scores when
  $n \geq p$) — which yield identical selection results;
- forward selection driven by a single omnibus **LRT** per iteration on
  the additive variance of the remaining composite term
  ($\Lambda \stackrel{\text{approx.}}{\sim} \chi^2_{1}$), which is
  already family-wise so **no Bonferroni correction is needed**;
- per-marker **outlier statistics**
  $o_k = \tilde{q}_k^2 / \tilde{v}_k$ used to nominate the next marker,
  subject to an LD-based `exclusion.window`;
- the final model
  $\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\tau}
    + \sum_{k=1}^{s} \hat{\beta}_k\,\boldsymbol{x}_k
    + \boldsymbol{Z}_u\boldsymbol{u}
    + \boldsymbol{Z}_g\boldsymbol{a}_{-s}
    + \boldsymbol{Z}_g\boldsymbol{p}
    + \boldsymbol{e}$
  in which the $s$ detected markers appear as fixed effects and the
  composite term is rebuilt on the remaining markers.

### What's different in GWAS

Only three things distinguish the GWAS pipeline from biparental interval
mapping at the modelling level:

1. **Marker matrix source.** `qtlAim()` operates on virtual
   mid-interval scores constructed by `primeCross(type = "interval")`
   ($\lambda$-weighted dosages in $[-1, +1]$). `gwasAim()` operates on
   the *actual* panel markers assembled by `primePanel()`, coded
   $\pm 1$ for the two alleles after the $x \mapsto x - 1$ shift from
   raw $0/1/2$ counts (heterozygotes at $0$). No interval mid-points
   are constructed; each panel marker is used directly.

2. **Population structure.** Diversity panels typically carry both
   short-range LD and long-range stratification from subpopulations.
   Both are absorbed by the composite additive term via
   $\boldsymbol{K}$ in the vm path, so no additional principal-
   component covariates or kinship correction are required in the
   base model.

3. **The vm path dominates.** Panels almost always satisfy $p \gg n$
   (many more markers than lines), so `gwasAim()` selects the vm
   formulation automatically. The mbf ridge formulation is available
   via `force = TRUE` when $n \geq p$ or for diagnostic comparison.

The **exclusion window** (argument `exclusion.window`, default 20) is
expressed in the units of the map's `pos` column — cM for a genetic
map, or Mb / kb / bp for a physical map.

---

## Data Preparation

The pre-analysis pipeline has three functions applied in sequence:

1. `checkPanel()` — pure diagnostic; reports data-quality issues but does
   not modify the data.
2. `filterPanel()` — applies quality thresholds in a principled order.
3. `primePanel()` — converts the clean matrix into a `wgPanel` object and
   optionally imputes any remaining missing values.

```{r load_data}
#| eval: !expr asreml_ok
data(genoGWASraw,  package = "wgAim")
data(mapGWASsim,   package = "wgAim")
data(phenoGWASsim, package = "wgAim")

cat(sprintf("Raw geno    : %d lines x %d marker columns\n",
            nrow(genoGWASraw), ncol(genoGWASraw) - 1L))
cat(sprintf("Map         : %d markers x %d columns\n",
            nrow(mapGWASsim), ncol(mapGWASsim)))
cat(sprintf("Phenotype   : %d rows across %d sites\n",
            nrow(phenoGWASsim), nlevels(phenoGWASsim$Site)))
```

A quick look at the raw genotype matrix — a `data.frame` with an `id`
column followed by one column per marker in 0/1/2 allele-count coding:

```{r peek_geno}
#| eval: !expr asreml_ok
genoGWASraw[1:6, 1:8]
```

### Step 1 — Diagnostic scan with `checkPanel()`

`checkPanel()` is a pure diagnostic. It computes summary statistics on
the raw matrix, cross-references marker names against the map, and reports
data-quality issues but does not modify the data. The results guide the
choice of thresholds for the subsequent `filterPanel()` step.

```{r checkpanel}
#| eval: !expr asreml_ok
chk <- checkPanel(genoGWASraw, mapGWASsim,
                   id       = "id",
                   map.id   = "marker",
                   map.chr  = "chr",
                   map.pos  = "pos",
                   encoding = "012")
print(chk)
```

The printed summary flags every embedded issue: 25 markers in the geno
matrix that are absent from the map (unplaceable), 25 markers in the map
absent from the geno matrix (the same markers under their original
names, before being renamed in the raw matrix), 4 duplicate lines,
roughly 40 duplicate marker pairs (the forced pairs plus a small number
of natural duplicates from short-range LD), 150 markers with excessive
missingness, 5 lines with excessive missingness, 15 monomorphic markers
and a MAF distribution consistent with a moderately dense panel. The MAF
threshold table makes it easy to preview how many markers would be
removed at each conventional cut-off.

### Step 2 — Apply filters with `filterPanel()`

`filterPanel()` uses the diagnostics to apply data-quality thresholds
in a principled order controlled by a `steps` argument. The default
`steps` list implements the standard workflow:

1. Map consistency — drop markers absent from the map.
2. Marker missingness (default 20%).
3. Line missingness (default 20%).
4. Duplicate lines.
5. Duplicate markers *(off by default; enable if desired)*.
6. MAF (default 0.05).

The `checkPanel` object can be piped directly into `filterPanel()`; all
metadata (id column, map field names, encoding) is carried through
automatically.

```{r filterpanel}
#| eval: !expr asreml_ok
flt <- filterPanel(chk)   # accepts the checkPanel object directly
print(flt)
```

The audit trail shows the effect of each step and confirms the final
panel dimensions (typically ~191 lines out of 200 and ~1450 markers out
of ~1700 with the default thresholds). Map consistency, the
high-missingness marker filter, monomorphic markers and duplicate
markers account for most of the marker attrition — the MAF filter is
effectively a no-op on this simulated panel since the raw genotypes are
already actively sampled at MAF > 0.15 (see the Overview). Line
attrition comes from the four seeded duplicate pairs plus five
high-missingness lines exceeding the 20% threshold.

### Step 3 — Build the `wgPanel` with `primePanel()`

`primePanel()` converts the filtered matrix into the internal object
class used by `gwasAim()`. It applies the $x \mapsto x-1$ encoding shift
so alleles are represented as $\pm 1$ (or dosage in $[-1, 1]$) with
heterozygotes at 0, splits the matrix by chromosome, sorts by position
within each chromosome and — optionally — imputes any remaining missing
values. The recommended in-package option is chromosome-wise KNN
imputation, which uses BLAS-accelerated pairwise distances and pre-sorted
neighbour indices to respect the LD structure within each chromosome.

```{r primepanel}
#| eval: !expr asreml_ok
panel <- primePanel(flt, impute = "knn", knn.k = 5)
cat("Class          :", paste(class(panel), collapse = ", "), "\n")
cat("Chromosomes    :", length(panel$geno), "\n")
cat("Markers total  :",
    sum(sapply(panel$geno, function(el) ncol(el$imputed.data))), "\n")
cat("Lines in panel :", nrow(panel$pheno), "\n")
```

The `wgPanel` object stores per-chromosome marker matrices in
`panel$geno[[ch]]$imputed.data` alongside their cM positions in
`panel$geno[[ch]]$map`.  A `linkMap()` call gives a visual summary of the
prepared panel:

```{r linkmap_panel}
#| eval: !expr asreml_ok
#| fig-height: 5
#| fig-width: 12
#| fig-cap: "Genetic map of the prepared diversity panel after filtering (10 chromosomes, ~170 markers each after quality filters)."
linkMap(panel, nrow = 1, marker.names = "none") +
  theme(plot.title = element_blank())
```

---

## Base Model — single environment (`Env1`)

For the single-environment GWAS we subset the phenotype to `Env1` and
fit a base model that captures the design structure. The trial is a
2-replicate RCBD with row and column plot coordinates. A simple and
robust choice is a random-effect model for lines, rows and columns:

```{r base_model}
#| eval: !expr asreml_ok
#| results: hide
env1 <- droplevels(phenoGWASsim[phenoGWASsim$Site == "Env1", ])
env1$Row    <- factor(env1$Row)
env1$Column <- factor(env1$Column)
env1$Rep    <- factor(env1$Rep)

base_env1 <- asreml(
  fixed     = yield ~ Rep,
  random    = ~ id + Row + Column,
  data      = env1,
  na.action = na.method(x = "include"),
  maxit     = 80
)
base_env1 <- update(base_env1)
```

The line term `id` captures line-to-line genetic variation prior to any
marker modelling; row and column random effects absorb spatial trend
across the field layout. Peek at the variance components to confirm all
terms are contributing:

```{r base_varcomp}
#| eval: !expr asreml_ok
summary(base_env1)$varcomp
```

---

## Running `gwasAim()`

`gwasAim()` takes the converged base model and the `wgPanel` object.
`merge.by` names the shared line-identifier column (present in both
`env1` and `panel$pheno`). The default `exclusion.window = 20` acts as
a proxy for LD-based exclusion — in this vignette the map is in cM,
so 20 means "no other detected marker within 20 cM":

```{r gwasaim}
#| eval: !expr asreml_ok
#| results: hide
gwas_env1 <- gwasAim(
  baseModel        = base_env1,
  genObj           = panel,
  merge.by         = "id",
  exclusion.window = 20,
  TypeI            = 0.05,
  trace            = FALSE,
  na.action        = na.method(x = "include"),
  maxit            = 80
)
```

The engine path (vm or mbf) is selected automatically. With ~191 lines
and ~1450 markers after filtering, we sit firmly in the vm regime and
the genomic relationship matrix
$\boldsymbol{K} = \tfrac{1}{p}\boldsymbol{M}\boldsymbol{M}^{\top}$ is used:

```{r gwas_class}
#| eval: !expr asreml_ok
#| results: markup
cat("Class              :", paste(class(gwas_env1), collapse = ", "), "\n")
cat("Markers in panel   :", gwas_env1$QTL$n.markers, "\n")
cat("Significant markers:", length(gwas_env1$QTL$qtl), "\n")
```

---

## Results

### `print()` — brief narrative

`print()` reports the significance threshold, panel size and one line
per detected marker with its chromosome, name and cM position:

```{r print_gwas}
#| eval: !expr asreml_ok
print(gwas_env1, genObj = panel)
```

### `summary()` — final marker table

`summary()` collates the final model into a table with one row per
detected marker: flanking marker name, cM position, estimated additive
effect, p-value, percentage of phenotypic variance explained and LOD
score:

```{r summary_gwas}
#| eval: !expr asreml_ok
summ_env1 <- summary(gwas_env1, genObj = panel)
print(summ_env1)
```

Rows are sorted by chromosome and cM position. The planted QTL (in
tonnes per hectare per allele substitution) are Chr2 +0.145 and Chr5
−0.145 (both MAIN); Chr7 +0.125 (weak MAIN); Chr3 fading from +0.230
in `Env1` down to 0 in `Env6`; and Chr9 crossing from −0.180 in
`Env1`–`Env3` to +0.180 in `Env4`–`Env6`. In a single environment
the strongest signals at `Env1` are the Chr3 fade (at its peak here)
and the Chr9 crossover (at its negative extreme), followed by the
two MAIN QTL on Chr2 and Chr5. The weakest MAIN (Chr7) may or may
not clear the threshold in a single environment — this is exactly the
kind of borderline effect that MV analysis is designed to detect by
borrowing strength across sites. See the roxygen help for
`?phenoGWASsim` for the shipped marker names and full QTL specification.

---

## Visualisation — `plot()`

Six complementary plot types are available via the `type` argument of
`plot.gwasAim()`. The default is `"manhattan"`, the classic GWAS display.

### Manhattan plot — `type = "manhattan"` (default)

A classic Manhattan plot: one point per marker, alternating chromosome
backgrounds, significant markers highlighted in `sig.col` with labels.
One facet per forward-selection iteration shows the outlier statistic
before each marker is nominated:

```{r plot_manhattan}
#| eval: !expr asreml_ok
#| fig-height: 9.6
#| fig-cap: "Manhattan plots across all forward-selection iterations. Each panel shows the genome-wide outlier statistic $\\tilde{q}_k^2/\\tilde{v}_k$; the significant marker for that iteration is highlighted and labelled."
plot(gwas_env1, genObj = panel, type = "manhattan", sig.col = "firebrick")
```

### Genome-wide outlier statistics — `type = "outlier"`

The outlier plot displays the same genome-wide statistic as line traces
faceted by iteration (an alternative to the Manhattan view). Chromosomal
regions that persistently carry signal show up as tall peaks across
multiple iterations:

```{r plot_outlier}
#| eval: !expr asreml_ok
#| fig-height: 9.6
#| fig-cap: "Line-plot view of the genome-wide outlier statistic at each forward-selection iteration. Vertical dashed lines mark the selected marker in each iteration."
plot(gwas_env1, genObj = panel, type = "outlier", sig.col = "firebrick")
```

### Scaled genome-wide BLUPs — `type = "blups"`

The BLUPs plot shows the scaled posterior means
$\tilde{q}_k / \sqrt{|\tilde{v}_k|}$ at every marker. Markers close to a
detected association naturally shrink towards zero as the association
absorbs the nearby signal in subsequent iterations:

```{r plot_blups}
#| eval: !expr asreml_ok
#| fig-height: 9.6
#| fig-cap: "Scaled composite BLUPs across the genome at each iteration."
plot(gwas_env1, genObj = panel, type = "blups", sig.col = "firebrick")
```

### Allele effects — `type = "effects"`

The effects lollipop summarises the final set of detected markers. Each
marker appears as a lollipop whose head sits at the estimated allele
substitution effect ± 1 SE, coloured by which allele is favoured:
**firebrick** when the +1 allele (raw genotype 2) increases yield,
**steelblue** when the −1 allele (raw genotype 0) does. The right panel
shows the percentage of variance explained per marker:

```{r plot_effects}
#| eval: !expr asreml_ok
#| fig-height: 5
#| fig-cap: "Allele effects and percentage variance explained for each detected marker."
plot(gwas_env1, genObj = panel, type = "effects")
```

### Allele contrast — `type = "contrast"`

The contrast plot displays the total genetic value for each line against
its dosage class (0, 1, 2) at each detected marker. The total genetic
value combines three components: the composite genome-wide BLUP, the
residual line-level genetic effect, and the summed fixed marker
contributions $\sum_k \hat{\beta}_k x_{ik}$.

This plot requires `data` — the augmented phenotypic data frame that
`gwasAim()` writes back to the calling environment as `<response>.data`
(here `yield.data`, which contains the original phenotype columns plus
`Gomit` and the `X.<chr>.<idx>` dosage columns for each detected marker):

```{r plot_contrast}
#| eval: !expr asreml_ok
#| fig-height: 7.2
#| fig-cap: "Total genetic value by dosage class (0/1/2) at each detected marker, faceted in two columns. Group means +/- SE are overlaid."
plot(gwas_env1, genObj = panel, type = "contrast",
     data = yield.data, ncol = 2)
```

### Iteration heatmap — `type = "heatmap"`

The heatmap gives a compact genome × iteration view, filling each cell
with the outlier statistic. This makes it easy to see which chromosomal
regions carried persistent signal across iterations:

```{r plot_heatmap}
#| eval: !expr asreml_ok
#| fig-height: 5
#| fig-cap: "Iteration × genome heatmap of the outlier statistic. Detected markers are annotated with a filled diamond."
plot(gwas_env1, genObj = panel, type = "heatmap", sig.col = "firebrick")
```

### Linkage map overlay — `linkMap()`

`linkMap()` overlays the detected markers on the panel map, with marker
names annotated on the right and the trait label on the left:

```{r linkmap_gwas}
#| eval: !expr asreml_ok
#| fig-height: 6
#| fig-width: 14
#| fig-cap: "Genetic map with detected GWAS markers overlaid. Marker names are annotated on the right; the trait label on the left."
linkMap(gwas_env1, genObj = panel, marker.names = "markers")
```

---

## Iteration Diagnostics — `aimTrace()`

`aimTrace()` traces how the LRT statistic and marker effect estimates
evolved across the forward-selection iterations. Two diagnostic plots
are available.

### LRT trace — `plot = "lrt"`

```{r aimtrace_lrt}
#| eval: !expr asreml_ok
#| fig-height: 4
#| fig-cap: "LRT statistic at each iteration. The dashed line is the chi-squared boundary threshold at TypeI = 0.05."
aimTrace(gwas_env1, plot = "lrt", print.out = FALSE)
```

### Effect stability — `plot = "stability"`

The stability plot traces each detected marker's allele effect ± 1 SE
across every iteration in which it was present in the model. Effects that
remain stable indicate the marker's contribution is well-identified and
not aliased with later-detected markers:

```{r aimtrace_stability}
#| eval: !expr asreml_ok
#| fig-height: 7
#| fig-cap: "Marker allele effect estimates +/- 1 SE across iterations, laid out 2 x 2."
aimTrace(gwas_env1, plot = "stability", print.out = FALSE, ncol = 2)
```

---

## Fine Mapping — `fineMap()`

`fineMap()` performs a high-resolution single-marker scan in a window
around each detected marker, holding all other detected effects fixed in
the model. For `gwasAim` objects this scans the *actual* panel markers
within `± window` cM of the peak (no interpolation), reporting per-marker
z-ratio p-values and LOD scores.

We fine-map the two markers detected first in forward selection — these
are the strongest signals by construction, since `gwasAim()` promotes
markers greedily by outlier statistic. A 25 cM window with the default
`exclusion.window = 20` gives good resolution around each peak:

```{r finemapping}
#| eval: !expr asreml_ok
#| results: hide
top2 <- head(gwas_env1$QTL$qtl, 2L)
fm <- fineMap(gwas_env1, genObj = panel,
              qtl              = top2,
              window           = 25,
              exclusion.window = 20)
```

The `plot()` method draws the LOD profile around each peak. The filled
triangle marks the peak fine-map position; the open triangle shows the
original coarse-scan position:

```{r finemap_plot}
#| eval: !expr asreml_ok
#| fig-height: 5
#| fig-width: 14
#| fig-cap: "Fine-mapping LOD profiles for the two strongest GWAS peaks."
plot(fm, ncol = 2)
```

---

## Summary

The complete single-environment GWAS pipeline illustrated in this
vignette is:

```r
# 1. Pre-analysis: check -> filter -> prime
chk   <- checkPanel(genoGWASraw, mapGWASsim, id = "id")
flt   <- filterPanel(chk)               # default step list
panel <- primePanel(flt, impute = "knn", knn.k = 5)

# 2. Subset phenotype to the environment of interest
env1  <- droplevels(phenoGWASsim[phenoGWASsim$Site == "Env1", ])
env1$Row    <- factor(env1$Row)
env1$Column <- factor(env1$Column)
env1$Rep    <- factor(env1$Rep)

# 3. Base model (no genomic term)
base_env1 <- asreml(yield ~ Rep, random = ~ id + Row + Column,
                    data = env1, na.action = na.method(x = "include"))

# 4. GWAS forward selection
gwas_env1 <- gwasAim(base_env1, genObj = panel, merge.by = "id",
                     TypeI = 0.05, exclusion.window = 20)

# 5. Results
summary(gwas_env1, genObj = panel)                # final marker table
plot(gwas_env1,    genObj = panel)                # Manhattan (default)
plot(gwas_env1,    genObj = panel, type = "outlier")
plot(gwas_env1,    genObj = panel, type = "blups")
plot(gwas_env1,    genObj = panel, type = "effects")
plot(gwas_env1,    genObj = panel, type = "contrast", data = yield.data)
plot(gwas_env1,    genObj = panel, type = "heatmap")
aimTrace(gwas_env1, plot = "lrt")
aimTrace(gwas_env1, plot = "stability")
linkMap(gwas_env1,  genObj = panel)

# 6. Fine mapping (top two markers in detection order)
top2 <- head(gwas_env1$QTL$qtl, 2L)
fm <- fineMap(gwas_env1, genObj = panel, qtl = top2, window = 25)
plot(fm, ncol = 2)
```

---

## References

Verbyla, A.P., Cullis, B.R. & Thompson, R. (2007). The analysis of QTL by
simultaneous use of the full linkage map. *Theoretical and Applied Genetics*,
**116**, 95–111. <https://doi.org/10.1007/s00122-007-0676-4>

Verbyla, A.P., Taylor, J.D. & Verbyla, K.L. (2012). RWGAIM: An efficient
high-dimensional random whole genome average (QTL) interval mapping approach.
*Genetics Research*, **94**, 291–306.
<https://doi.org/10.1017/S0016672312000493>

Taylor, J. & Verbyla, A.P. (2011). R Package wgaim: QTL Analysis in
Bi-Parental Populations Using Linear Mixed Models. *Journal of Statistical
Software*, **40**(7), 1–18. <https://doi.org/10.18637/jss.v040.i07>

VanRaden, P.M. (2008). Efficient methods to compute genomic predictions.
*Journal of Dairy Science*, **91**(11), 4414–4423.
<https://doi.org/10.3168/jds.2007-0980>
