---
title: "Multi-Environment GWAS with wgAim"
author: "Julian Taylor & Ari Verbyla"
date: today
vignette: >
  %\VignetteIndexEntry{Multi-Environment GWAS with wgAim}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 10
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
  cache: true
---

```{r setup}
#| include: false
#| cache: false
asreml_ok <- requireNamespace("asreml", quietly = TRUE)
if (requireNamespace("wgAim", quietly = TRUE)) {
  library(wgAim)
} else {
  devtools::load_all("C:/jtaylor/Research/Computing/R Packages/Claude/wgaim",
                     quiet = TRUE)
}
library(ggplot2)
if (asreml_ok) {
  library(asreml)
  asreml.options(trace = FALSE)
}
```

---

## Overview

This vignette demonstrates the multi-environment extension of the **wgAim**
GWAS pipeline for a diversity panel evaluated across several field trials.
The framework detects markers whose effects are consistent across trials
(**main-effect QTL**, MAIN) as well as markers whose effects differ between
trials (**G×E interaction QTL**, INT), within a unified forward-selection
model — exactly as the multivariate `qtlAim()` engine does for biparental
crosses.

The pipeline has the same three stages as the univariate GWAS analysis:

1. **Pre-analysis** — check, filter and prime the diversity panel with
   `checkPanel()`, `filterPanel()` and `primePanel()`.
2. **Analysis** — develop a multi-trial base model and run `gwasAim()`
   with the `Trait` argument set to the grouping factor (here `"Site"`).
3. **Post-analysis** — interrogate, summarise, and visualise the detected
   markers using `summary()`, `plot()`, and `aimTrace()`.

The `Trait` argument is a general grouping factor: it may represent sites,
trials, environments, seasons, or measured traits — any factor whose
levels define the multivariate response. In this vignette `Trait = "Site"`.

Throughout we use the `genoGWASraw`, `mapGWASsim` and `phenoGWASsim`
datasets bundled with the package. `genoGWASraw` and `mapGWASsim`
represent a simulated diversity panel of 200
lines genotyped at approximately 1700 SNP markers across 10
chromosomes (~170 markers per chromosome, average spacing ~0.9 cM),
evaluated in 6 field sites with 2 replicates each. The panel is
generated with `AlphaSimR` with 50 founder haplotypes via a coalescent
(`runMacs2`), one generation of random mating to produce 200 F1
diverse lines, then a SNP chip actively selected from segregating
sites with minor allele frequency (MAF) > 0.15. All shipped markers
have F1 MAF > 0.15 (median ~0.32), giving a kinship matrix
$\boldsymbol{K}$ with genuine full-rank structure (off-diagonal SD
~0.085) that captures the half-sib / full-sib relatedness naturally.
Yields are on the tonnes-per-hectare scale (site means 4.5–4.9 t/ha,
residual SD ~0.377 t/ha per site) with two structured components
matching the multivariate analysis's
`fa(Site,k):vm(id, K) + fa(Site,k):id` decomposition (both kept
deliberately small so the composite LRT stops cleanly once true QTL
are absorbed). Five QTL were planted at known positions to give ground
truth for detection: three with consistent effects across all sites
(Chr2 +0.145 t/ha, Chr5 −0.145 t/ha, and Chr7 +0.125 t/ha — the
third being deliberately weak to sit near the detection threshold),
and two with site-specific effects (Chr3 fading from +0.230 t/ha in
`Env1` to 0 in `Env6`; Chr9 crossing from −0.180 t/ha in `Env1`–`Env3`
to +0.180 t/ha in `Env4`–`Env6`). The raw marker matrix contains
realistic data-quality issues (missingness, monomorphic markers,
duplicate lines, duplicate markers, unplaceable markers) so the full
pre-analysis pipeline can be exercised. See the roxygen help for
`?phenoGWASsim` for the full specification.

::: {.note}
**Requires:** ASReml-R V4. The vignette is pre-rendered so no licence is
needed to read it. Run the code chunks only if you have ASReml-R installed.
:::

---

## Mathematical Framework

The mathematical framework underpinning multi-environment `gwasAim()` is
essentially identical to that of multivariate `qtlAim()`. Please see the
**Mathematical Framework** section of the [Multivariate QTL Mapping
vignette](qtlAim_mv_pipeline.html#mathematical-framework) for the full algorithmic pipeline.

### What's different in GWAS

The same three GWAS-specific differences noted in the univariate GWAS
vignette apply here:


1. **Marker matrix source.** `qtlAim()` uses virtual mid-interval scores
   from `primeCross(type = "interval")`. `gwasAim()` uses the *actual*
   panel markers assembled by `primePanel()`, coded $\pm 1$ after the
   $x \mapsto x - 1$ shift from raw $0/1/2$ counts (heterozygotes at
   $0$). Each panel marker is used directly.

2. **Population structure.** Diversity panels typically carry both
   short-range LD and long-range stratification from subpopulations;
   both are absorbed by the composite additive term via
   $\boldsymbol{G}_a \otimes \boldsymbol{K}$ in the vm path.

3. **The vm path dominates.** Panels almost always satisfy $p \gg n$,
   so the multivariate vm formulation
   $\boldsymbol{a} \sim \mathcal{N}(\boldsymbol{0},\, \boldsymbol{G}_a \otimes \boldsymbol{K})$
   with $\boldsymbol{K} = \tfrac{1}{p}\boldsymbol{M}\boldsymbol{M}^{\top}$
   is selected automatically.

---

## Data Preparation

The pre-analysis pipeline is identical to the univariate GWAS vignette:
`checkPanel()` for a diagnostic scan, `filterPanel()` for quality
filtering with a principled step order, and `primePanel()` for
$\pm 1$ recoding, chromosome-wise splitting and residual KNN imputation.
See the [Univariate GWAS vignette](gwasAim_pipeline.html#data-preparation)
for detailed narrative on each function. Here we run all three in a single
condensed block:

```{r panel_prep}
#| eval: !expr asreml_ok
data(genoGWASraw,  package = "wgAim")
data(mapGWASsim,   package = "wgAim")
data(phenoGWASsim, package = "wgAim")

chk   <- checkPanel(genoGWASraw, mapGWASsim, id = "id")
flt   <- filterPanel(chk)
panel <- primePanel(flt, impute = "knn", knn.k = 5)

cat(sprintf("Prepared panel : %d lines x %d markers across %d chromosomes\n",
            nrow(panel$pheno),
            sum(sapply(panel$geno, function(el) ncol(el$imputed.data))),
            length(panel$geno)))
```

With ~191 lines and ~1500 markers after filtering, the panel sits firmly
in the $p \gg n$ regime and `gwasAim()` will use the vm path automatically
at every iteration of forward selection.

### Phenotype setup

The phenotype is a balanced 6-site, 2-replicate MET with `Site`, `id`
and `Rep` as factors. We use *all* 6 sites in the multi-environment
analysis (the univariate GWAS vignette used only `Env1`). The site names
are zero-padded to the minimum width needed — for 6 sites this gives
`Env1`..`Env6`:

```{r pheno_setup}
#| eval: !expr asreml_ok
pheno <- phenoGWASsim
pheno$id   <- factor(pheno$id)
n.sites    <- nlevels(pheno$Site)
site.width <- nchar(as.character(n.sites))
env.names  <- sprintf(paste0("Env%0", site.width, "d"), seq_len(n.sites))
pheno$Site <- factor(pheno$Site, levels = env.names)
pheno$Rep  <- factor(pheno$Rep)

cat(sprintf("Phenotype : %d rows across %d sites (%d lines x %d reps per site)\n",
            nrow(pheno), nlevels(pheno$Site),
            nlevels(pheno$id), nlevels(pheno$Rep)))
```

---

## Base Model Development

The starting point for a multi-environment `gwasAim()` analysis is a
converged multi-trial base model that captures the experimental design
structure and the genetic (co)variance across sites. We build this
incrementally, comparing three variance structures for the genetic term.

### Step 1 — Diagonal structure

The simplest multi-site model assumes genetic effects in each site are
**independent**:
$\boldsymbol{G}_a = \text{diag}(\sigma^2_{g,1}, \ldots, \sigma^2_{g,T})$
with no between-site genetic correlations.

```{r base_diag}
#| eval: !expr asreml_ok
#| results: hide
base_diag <- asreml(
  fixed     = yield ~ Site + Site:Rep,
  random    = ~ diag(Site):id,
  residual  = ~ dsum(~units | Site),
  data      = pheno,
  na.action = na.method(x = "include"),
  maxit     = 100
)
base_diag <- update(base_diag)
```

### Step 2 — Factor analytic order 1

A single-factor model FA(1) introduces one set of loadings and captures
the dominant axis of genotype-by-site interaction:

```{r base_fa1}
#| eval: !expr asreml_ok
#| results: hide
base_fa1 <- asreml(
  fixed     = yield ~ Site + Site:Rep,
  random    = ~ fa(Site, 1):id,
  residual  = ~ dsum(~units | Site),
  data      = pheno,
  na.action = na.method(x = "include"),
  maxit     = 100
)
base_fa1 <- update(base_fa1)
```

### Step 3 — Factor analytic order 2

A two-factor model FA(2) captures two axes of interaction and provides
a richer description of the genetic covariance. This is the model we
carry into `gwasAim()`:

```{r base_fa2}
#| eval: !expr asreml_ok
#| results: hide
base_fa2 <- asreml(
  fixed     = yield ~ Site + Site:Rep,
  random    = ~ fa(Site, 2):id,
  residual  = ~ dsum(~units | Site),
  data      = pheno,
  na.action = na.method(x = "include"),
  maxit     = 100
)
base_fa2 <- update(base_fa2)
```

### Model comparison

We compare the three variance structures using their restricted
log-likelihoods and an AIC-style penalty. Parameter counts for `n.sites`
= 6 sites are: `diag` = 6 (variance per site); `fa1` = 6 psi + 6 loadings
= 12; `fa2` = 6 psi + 12 loadings = 18. Each model also carries the
6 residual variances from `dsum(~units | Site)`:

```{r model_comparison}
#| eval: !expr asreml_ok
np <- c(diag = as.integer(   n.sites + n.sites),
        fa1  = as.integer(2L*n.sites + n.sites),
        fa2  = as.integer(3L*n.sites + n.sites))
comp <- data.frame(
  Model      = c("diag(Site):id", "fa(Site, 1):id", "fa(Site, 2):id"),
  Parameters = np,
  LogLik     = round(c(base_diag$loglik, base_fa1$loglik, base_fa2$loglik), 2),
  AIC        = round(c(-2*base_diag$loglik + 2*np["diag"],
                        -2*base_fa1$loglik  + 2*np["fa1"],
                        -2*base_fa2$loglik  + 2*np["fa2"]),   2),
  stringsAsFactors = FALSE
)
print(comp, row.names = FALSE)
```

FA(2) provides the best balance between fit and parsimony while capturing
two axes of genotype-by-site interaction across the six sites.

---

## GWAS Mapping — `gwasAim()`

### Running `gwasAim()`

The `Trait` argument activates the multivariate engine, treating
`"Site"` as the grouping factor. Leaving `str = NULL` (the default)
instructs `gwasAim()` to **mirror** the variance structure of the base
model — here FA(2) — for the composite genome-wide random term, giving
consistent modelling of the genetic covariance at every iteration
without needing to respecify the order.

We use the default significance level `TypeI = 0.05`, applied at the
family-wise level (the single omnibus multivariate LRT per iteration
does not need Bonferroni correction — see the referenced Mathematical
Framework section for the argument):

```{r gwasAim}
#| eval: !expr asreml_ok
#| results: hide
gwas_mv <- gwasAim(
  baseModel        = base_fa2,
  genObj           = panel,
  merge.by         = "id",
  Trait            = "Site",
  str              = NULL,            # mirror the FA(2) structure
  exclusion.window = 20,
  TypeI            = 0.05,
  trace            = FALSE,
  na.action        = na.method(x = "include"),
  maxit            = 100
)
```

```{r gwasAim_class}
#| eval: !expr asreml_ok
#| results: markup
cat("Class              :", paste(class(gwas_mv), collapse = ", "), "\n")
cat("Trait              :", gwas_mv$QTL$Trait, "\n")
cat("Markers in panel   :", gwas_mv$QTL$n.markers, "\n")
cat("Significant markers:", length(gwas_mv$QTL$qtl), "\n")
```

---

## Results

### `print()` — brief narrative

`print()` reports the significance threshold, panel size and one line
per detected marker with its chromosome, name and cM position, plus a
`[MAIN]` or `[INTERACTION]` tag from the post-selection Wald pruning:

```{r print_gwas}
#| eval: !expr asreml_ok
print(gwas_mv, genObj = panel)
```

### `summary()` — final marker table

The multivariate `summary()` returns a data frame whose row structure
depends on each detected marker's MAIN vs G×E classification from the
post-selection Wald pruning:

- **MAIN markers** produce a **single row** with `Trait = "MAIN"` and
  a single `Size` / `Pvalue` / `LOD` value (one common effect across
  all sites).
- **G×E markers** produce **one row per site**, with `Trait` set to
  the site level (`Env1`..`Env6`) and per-site `Size` / `Pvalue` /
  `LOD` values.

The `Perc.Var` column is computed once per unique QTL and broadcast
across its rows (so all six rows of a G×E QTL carry the same
`Perc.Var`). Rows are sorted by chromosome, then cM position, then
`Trait`:

```{r summary_gwas}
#| eval: !expr asreml_ok
summ_mv <- summary(gwas_mv, genObj = panel)
print(summ_mv)
```

The multi-environment analysis borrows strength across the six sites,
so weaker per-site signals combine into detectable joint effects.
Compare the recovery here to the single-environment analysis in
`Env1` from the [Univariate GWAS
vignette](gwasAim_pipeline.html): the clearest illustration of the
MV advantage in this dataset is the recovery of the **Chr5 MAIN**
QTL (−0.145 t/ha), which sits below the detection threshold in the
single-environment analysis but is comfortably resolved here once
the LRT combines evidence across all six sites. The G×E QTL are
also more informatively described in the MV analysis: Chr9's
crossover (±0.180 t/ha) is correctly classified as an INTERACTION
with a sign flip visible in the per-site `Size` estimates, and
Chr3's fade shows large positive per-site effects in `Env1`–`Env4`
decaying towards zero in `Env5`–`Env6`, exactly matching its
planted +0.230 → 0 trajectory.

---

## Visualisation — `plot()`

Six complementary plot types are available via the `type` argument of
`plot.gwasAim()`. The default for a multivariate object is
`"manhattan"`.

### Manhattan plot — `type = "manhattan"` (default)

A classic Manhattan plot: one point per marker, alternating chromosome
backgrounds, significant markers highlighted in `sig.col` with labels.
For a multivariate object the y-axis is the quadratic-form outlier
statistic $o_k$, and one facet is drawn per forward-selection iteration:

```{r plot_manhattan}
#| eval: !expr asreml_ok
#| fig-height: 9.6
#| fig-cap: "Manhattan plots across all forward-selection iterations. Each panel shows the multivariate quadratic-form outlier statistic $o_k = \\tilde{\\boldsymbol{q}}_k^\\top \\boldsymbol{G}_a^{-1} \\tilde{\\boldsymbol{q}}_k / \\tilde{v}_k$; the significant marker for that iteration is highlighted and labelled."
plot(gwas_mv, genObj = panel, type = "manhattan", sig.col = "firebrick")
```

### Genome-wide outlier statistics — `type = "outlier"`

The outlier plot displays the same multivariate quadratic-form
statistic as line traces faceted by iteration. Chromosomal regions
that persistently carry signal show up as tall peaks across multiple
iterations:

```{r plot_outlier}
#| eval: !expr asreml_ok
#| fig-height: 9.6
#| fig-cap: "Line-plot view of the multivariate outlier statistic at each forward-selection iteration. Vertical dashed lines mark the selected marker in each iteration."
plot(gwas_mv, genObj = panel, type = "outlier", sig.col = "firebrick")
```

### Scaled genome-wide BLUPs — `type = "blups"`

For a multivariate GWAS object the BLUPs plot shows **one line per
site**, coloured by site. This display is particularly informative:
markers where all site lines move together indicate main-effect
signal, while markers where site lines diverge or cross indicate G×E
signal (in the raw scaled-BLUP domain, before the model has committed
to MAIN vs INT classification):

```{r plot_blups}
#| eval: !expr asreml_ok
#| fig-height: 9.6
#| fig-cap: "Scaled composite BLUPs across the genome at each iteration, one line per site. Convergent lines indicate consistent main-effect signal; divergent or crossing lines indicate G×E signal."
plot(gwas_mv, genObj = panel, type = "blups", sig.col = "firebrick")
```

### Allele effects — `type = "effects"`

The effects lollipop is extended to show one row per site per marker.
Main-effect QTL appear as a tight cluster of identically coloured
lollipops across all sites; G×E QTL show variation in magnitude or
sign across sites:

```{r plot_effects}
#| eval: !expr asreml_ok
#| fig-height: 7
#| fig-cap: "Per-site allele effects ± SE for each detected marker. Firebrick = +1 allele (raw genotype 2) favoured; steelblue = −1 allele (raw genotype 0) favoured. Main-effect markers show consistent colours across sites; G×E markers show site-dependent variation."
plot(gwas_mv, genObj = panel, type = "effects")
```

### Allele contrast — `type = "contrast"`

The contrast plot shows the total genetic value per line against its
allele dosage class (0, 1, 2) at each detected marker, separately for
each site. Main-effect markers show the same direction of separation
across all site panels; G×E markers show the separation varying or
reversing between site panels. As in the univariate vignette, this
plot requires `data` — the augmented phenotype data frame that
`gwasAim()` writes back to the calling environment as `yield.data`:

```{r plot_contrast}
#| eval: !expr asreml_ok
#| fig-height: 12
#| fig-cap: "Total genetic value by allele dosage class (0/1/2) at each detected marker, faceted by site (columns) and marker (rows). Group means ± SE overlaid."
plot(gwas_mv, genObj = panel, type = "contrast",
     data = yield.data, ncol = 3)
```

### Iteration heatmap — `type = "heatmap"`

The heatmap gives a compact genome × iteration view, filling each cell
with the multivariate outlier statistic. This makes it easy to see
which chromosomal regions carried persistent multi-environment signal
across iterations:

```{r plot_heatmap}
#| eval: !expr asreml_ok
#| fig-height: 5
#| fig-cap: "Iteration × genome heatmap of the multivariate outlier statistic. Detected markers are annotated."
plot(gwas_mv, genObj = panel, type = "heatmap", sig.col = "firebrick")
```

### Linkage map overlay — `linkMap()`

`linkMap()` overlays the detected markers on the panel map:

```{r linkmap_gwas}
#| eval: !expr asreml_ok
#| fig-height: 6
#| fig-width: 14
#| fig-cap: "Genetic map with detected multi-environment GWAS markers overlaid. Marker names annotated on the right; the trait label on the left."
linkMap(gwas_mv, genObj = panel, marker.names = "markers")
```

---

## Iteration Diagnostics — `aimTrace()`

### LRT trace — `plot = "lrt"`

The mixture chi-squared threshold (shown as a dashed line) reflects
the $T$-dimensional nature of the test — it lies above the univariate
$\chi^2_1$ threshold because the test jointly evaluates effects across
all $T = 6$ sites:

```{r aimtrace_lrt}
#| eval: !expr asreml_ok
#| fig-height: 4
#| fig-cap: "LRT statistic at each forward-selection iteration. The dashed line is the chi-squared mixture threshold at TypeI = 0.05 for T = 6 sites."
aimTrace(gwas_mv, plot = "lrt", print.out = FALSE)
```

### Effect stability — `plot = "stability"`

The stability plot traces the per-site allele effect ± 1 SE for each
detected marker across iterations. Stable main-effect markers show all
site lines tracking closely together; G×E markers show site lines that
are separated or diverge as further markers enter the model. The panel
is laid out with sites as rows and markers as columns, with a
`[MAIN]` or `[INT]` tag in each column header:

```{r aimtrace_stability}
#| eval: !expr asreml_ok
#| fig-height: 9.8
#| fig-cap: "Per-site allele effect estimates ± 1 SE across iterations. Rows = sites; columns = detected markers with [MAIN] or [INT] suffix. Stable markers show consistent effects across iterations; G×E markers show site-dependent variation."
aimTrace(gwas_mv, plot = "stability", print.out = FALSE)
```

---

## Summary

The complete multi-environment GWAS pipeline illustrated in this
vignette is:

```r
# 1. Pre-analysis: check -> filter -> prime the diversity panel
chk   <- checkPanel(genoGWASraw, mapGWASsim, id = "id")
flt   <- filterPanel(chk)
panel <- primePanel(flt, impute = "knn", knn.k = 5)

# 2. Phenotype setup
pheno       <- phenoGWASsim
pheno$id    <- factor(pheno$id)
pheno$Site  <- factor(pheno$Site)   # ensures Env1..Env6 ordering
pheno$Rep   <- factor(pheno$Rep)

# 3. Multi-trial base model progression (diag -> FA1 -> FA2)
base_diag <- asreml(yield ~ Site + Site:Rep,
                    random   = ~ diag(Site):id,
                    residual = ~ dsum(~units | Site),
                    data = pheno, na.action = na.method(x = "include"))
base_fa1  <- asreml(yield ~ Site + Site:Rep,
                    random   = ~ fa(Site, 1):id,
                    residual = ~ dsum(~units | Site),
                    data = pheno, na.action = na.method(x = "include"))
base_fa2  <- asreml(yield ~ Site + Site:Rep,
                    random   = ~ fa(Site, 2):id,
                    residual = ~ dsum(~units | Site),
                    data = pheno, na.action = na.method(x = "include"))

# 4. Multi-environment GWAS forward selection (str = NULL mirrors FA(2))
gwas_mv <- gwasAim(base_fa2, genObj = panel, merge.by = "id",
                   Trait = "Site", str = NULL,
                   TypeI = 0.05, exclusion.window = 20,
                   na.action = na.method(x = "include"), maxit = 100)

# 5. Results
summary(gwas_mv, genObj = panel)                 # MAIN/INT table by site
plot(gwas_mv,    genObj = panel)                 # Manhattan (default)
plot(gwas_mv,    genObj = panel, type = "outlier")
plot(gwas_mv,    genObj = panel, type = "blups")
plot(gwas_mv,    genObj = panel, type = "effects")
plot(gwas_mv,    genObj = panel, type = "contrast",
                 data = yield.data, ncol = 3)
plot(gwas_mv,    genObj = panel, type = "heatmap")
aimTrace(gwas_mv, plot = "lrt",       print.out = FALSE)
aimTrace(gwas_mv, plot = "stability", print.out = FALSE)
linkMap(gwas_mv,  genObj = panel)
```

---

## References

Verbyla, A.P., Cullis, B.R. & Thompson, R. (2007). The analysis of QTL by
simultaneous use of the full linkage map. *Theoretical and Applied Genetics*,
**116**, 95–111. <https://doi.org/10.1007/s00122-007-0676-4>

Verbyla, A.P., Taylor, J.D. & Verbyla, K.L. (2012). RWGAIM: An efficient
high-dimensional random whole genome average (QTL) interval mapping approach.
*Genetics Research*, **94**, 291–306.
<https://doi.org/10.1017/S0016672312000493>

Taylor, J. & Verbyla, A.P. (2011). R Package wgaim: QTL Analysis in
Bi-Parental Populations Using Linear Mixed Models. *Journal of Statistical
Software*, **40**(7), 1–18. <https://doi.org/10.18637/jss.v040.i07>

Smith, A.B., Cullis, B.R. & Thompson, R. (2001). Analyzing variety by
environment data using multiplicative mixed models and adjustments for
spatial field trend. *Biometrics*, **57**(4), 1138–1147.
<https://doi.org/10.1111/j.0006-341X.2001.01138.x>

VanRaden, P.M. (2008). Efficient methods to compute genomic predictions.
*Journal of Dairy Science*, **91**(11), 4414–4423.
<https://doi.org/10.3168/jds.2007-0980>
