---
title: "Multivariate QTL Mapping with wgAim"
author: "Julian Taylor & Ari Verbyla"
date: today
vignette: >
  %\VignetteIndexEntry{Multivariate QTL Mapping with wgAim}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 10
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
  cache: true
---

```{r setup}
#| include: false
#| cache: false
asreml_ok <- requireNamespace("asreml", quietly = TRUE)
if (requireNamespace("wgAim", quietly = TRUE)) {
  library(wgAim)
} else {
  devtools::load_all("C:/jtaylor/Research/Computing/R Packages/Claude/wgaim",
                     quiet = TRUE)
}
library(ggplot2)
if (asreml_ok) {
  library(asreml)
  asreml.options(trace = FALSE)
}
```

---

## Overview

This vignette demonstrates the multivariate extension of the **wgAim** QTL
mapping pipeline for a biparental mapping population evaluated across multiple
environments. The framework detects QTL that act consistently across
environments (**main-effect QTL**) as well as QTL whose effects differ
between environments (**G×E interaction QTL**), within a unified
forward-selection model.

The pipeline has the same three stages as the univariate analysis:

1. **Pre-analysis** — prepare the `wgCross` object using `primeCross()`.
2. **Analysis** — develop a multi-environment base model and run
   `qtlAim()` with the `Trait` argument to activate the multivariate engine.
3. **Post-analysis** — interrogate, summarise, and visualise the detected
   QTL using `summary()`, `plot()`, and `aimTrace()`.

Throughout we use the `genoMVsim` and `phenoMVsim` datasets bundled with the
package: a simulated Doubled Haploid (DH) population of 120 lines
genotyped at 600 markers across 6 chromosomes, evaluated in 8 field trials
with 2 replicates each. Four QTL were planted: two with consistent effects
across trials (main effect) and two with environment-specific effects (G×E).

::: {.note}
**Requires:** ASReml-R V4. The vignette is pre-rendered so no licence is
needed to read it. Run the code chunks only if you have ASReml-R installed.
:::

---

## Mathematical Framework

### The multivariate linear mixed model

Let $t = 1, \ldots, T$ index environments (trials) and let
$\boldsymbol{y}_t$ be the $n \times 1$ vector of phenotypic observations
for environment $t$. The stacked response across all environments is
$\boldsymbol{y} = (\boldsymbol{y}_1^\top, \ldots, \boldsymbol{y}_T^\top)^\top$.
The base linear mixed model is

$$
\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\tau}
               + \boldsymbol{Z}_u\boldsymbol{u}
               + \boldsymbol{Z}_g\boldsymbol{g}
               + \boldsymbol{e},
$$

where $\boldsymbol{X}$ and $\boldsymbol{Z}_u$ are design matrices for the
fixed effects $\boldsymbol{\tau}$ (environment means, replicate effects) and
peripheral random effects $\boldsymbol{u}$ respectively, and
$\boldsymbol{e} \sim \mathcal{N}(\boldsymbol{0},\, \boldsymbol{\Sigma}_e)$
with $\boldsymbol{\Sigma}_e$ a block-diagonal residual structure allowing
environment-specific error variances. The genetic line effects
$\boldsymbol{g}$ are modelled jointly across environments as

$$
\boldsymbol{g} \sim \mathcal{N}\!\left(\boldsymbol{0},\,
\boldsymbol{G}_a \otimes \boldsymbol{I}_n\right),
$$

where $\boldsymbol{G}_a$ is the $T \times T$ genetic (co)variance matrix
capturing the magnitude of genetic variation in each environment and the
genetic correlations between environments.

### Factor analytic structure for $\boldsymbol{G}_a$

With many environments, $\boldsymbol{G}_a$ has $T(T+1)/2$ free parameters
— too many to estimate reliably unless the number of lines is large. A
parsimonious and biologically interpretable approximation is the
**factor analytic** (FA) model of order $k$:

$$
\boldsymbol{G}_a = \boldsymbol{\Lambda}\boldsymbol{\Lambda}^\top
                 + \boldsymbol{\Psi},
$$

where $\boldsymbol{\Lambda}$ is the $T \times k$ matrix of factor
**loadings** and $\boldsymbol{\Psi} = \text{diag}(\psi_1, \ldots, \psi_T)$
is a diagonal matrix of environment-specific **specific variances**. The
loadings capture the dominant axes of genotype-by-environment interaction,
while the specific variances absorb any residual environment-specific
genetic variation not explained by the common factors. An FA($k$) model
has $Tk + T = T(k+1)$ parameters, which is much smaller than the full
$T(T+1)/2$ when $k \ll T$.

### Introducing additive genetic effects

As in the univariate case, the genetic term $\boldsymbol{g}$ is decomposed
into an **additive composite term** $\boldsymbol{a}$ — representing the
cumulative effect of all markers across the genome — and a **residual
polygenic term** $\boldsymbol{p}$ that absorbs genetic signal not captured
by any detected QTL:

$$
\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\tau}
               + \boldsymbol{Z}_u\boldsymbol{u}
               + \boldsymbol{Z}_g\boldsymbol{a}
               + \boldsymbol{Z}_g\boldsymbol{p}
               + \boldsymbol{e}.
$$

The composite additive term uses one of two equivalent computational paths
depending on the ratio of marker intervals $p$ to lines $n$:

- **vm path** ($p > n$): The aggregate marker signal is absorbed into a
  genomic relationship matrix
  $\boldsymbol{K} = \tfrac{1}{p}\boldsymbol{M}\boldsymbol{M}^{\top}$,
  and
  $\boldsymbol{a} \sim \mathcal{N}(\boldsymbol{0},\, \boldsymbol{G}_a \otimes \boldsymbol{K})$.
  This is the multivariate extension of the standard GBLUP formulation,
  efficient when there are more intervals than lines.
- **mbf path** ($n \geq p$): Each marker effect is treated as an independent
  random regression coefficient,
  $\boldsymbol{a} = (\boldsymbol{I}_T \otimes \boldsymbol{M})\boldsymbol{q}$
  with
  $\boldsymbol{q} \sim \mathcal{N}(\boldsymbol{0},\, \boldsymbol{G}_a \otimes \sigma^2_q\,\boldsymbol{I}_p)$.
  This path is preferred when lines outnumber intervals.

Both paths yield identical QTL selection results. The engine selects the
appropriate path automatically; the `force = TRUE` argument overrides to
the mbf path. In this vignette, with 594 intervals and 120 lines ($p > n$),
the vm path is used throughout, and $\boldsymbol{G}_a$ carries the same
factor analytic structure as the base model.

### Forward selection and the likelihood ratio test

Forward selection proceeds identically to the univariate case. At each
iteration a likelihood ratio test (LRT) is used to determine whether any
interval or marker carries significant additive signal. For $T$ environments,
however, each candidate is tested as a $T$-vector of fixed effects — one
per environment — so the test statistic follows a **chi-squared mixture**
distribution rather than a standard $\chi^2_1$:

$$
\Lambda = -2\!\left(\ell_{\text{reduced}} - \ell_{\text{full}}\right)
        \;\stackrel{\text{approx.}}{\sim}\;
        \sum_{j=0}^{T} w_j \,\chi^2_j,
$$

where the mixing weights $w_j = \binom{T}{j} \tfrac{1}{2^T}$ arise from the
boundary constraints on variance components, and $\ell$ is the restricted
log-likelihood. The mixture critical value at family-wise significance level
`TypeI` is obtained numerically via `qchisq.mixture()`.

### Genome-wide outlier statistic

Because the BLUP of interval $k$'s additive effect is now a $T$-vector
$\tilde{\boldsymbol{q}}_k$, the outlier statistic is a **quadratic form**
in $\boldsymbol{G}_a^{-1}$:

$$
o_k = \frac{\tilde{\boldsymbol{q}}_k^\top\,
       \boldsymbol{G}_a^{-1}\,\tilde{\boldsymbol{q}}_k}{\tilde{v}_k},
$$

where the normalising scalar
$\tilde{v}_k = \text{tr}\!\left(\boldsymbol{G}_a^{-1}\,
\boldsymbol{T}_k\,\tilde{\boldsymbol{V}}_{\boldsymbol{a}}\,
\boldsymbol{T}_k^\top\right)$,
$\boldsymbol{T}_k = \boldsymbol{I}_T \otimes \boldsymbol{t}_k^\top$ is the
extraction matrix for interval $k$, and
$\tilde{\boldsymbol{V}}_{\boldsymbol{a}}$ is the posterior variance matrix
of the composite random term. This statistic is non-negative and
approximately chi-squared scaled, placing all intervals on a common scale
regardless of the number of environments.

### Post-selection: main-effect vs G×E QTL

After all QTL have been selected, a Wald zero-equality test is applied to the
$T$-vector of fixed effects for each detected QTL. QTL whose effects are
consistent across environments (main effects) are retained with a single fixed
coefficient; those with significant environment-specific variation are retained
as a full $T$-vector of environment-specific effects (G×E interaction QTL).
The final model is

$$
\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\tau}
               + \boldsymbol{Z}_g\!\sum_{j \in \mathcal{M}} \hat{\beta}_j\,\boldsymbol{x}_j
               + \boldsymbol{Z}_g\!\sum_{k \in \mathcal{I}}
                 \hat{\boldsymbol{\gamma}}_k^\top\,\boldsymbol{x}_k
               + \boldsymbol{Z}_u\boldsymbol{u}
	       + \boldsymbol{Z}_g\boldsymbol{a}_{-s}
               + \boldsymbol{Z}_g\boldsymbol{p}
               + \boldsymbol{e},
$$

where  $\boldsymbol{a}_{-s} = \boldsymbol{M}_{-s}\boldsymbol{q}_{-s}$
where the $\boldsymbol{M}_{-s}$ denotes $\boldsymbol{M}$ with an omission of
markers/intervals in the windows of the QTL selected; $\mathcal{M}$ denotes
main-effect QTL with scalar effects $\hat{\beta}_j and $\mathcal{I}$ denotes
G×E QTL with environment-specific effect vectors $\hat{\boldsymbol{\gamma}}_k$.

---

## Data Preparation — `primeCross()`

```{r load_data}
#| eval: !expr asreml_ok
data(genoMVsim,  package = "wgAim")
data(phenoMVsim, package = "wgAim")
```

`primeCross()` converts the `cross` object into a `wgCross` object with
interval genotype scores ready for `qtlAim()`:

```{r primeCross}
#| eval: !expr asreml_ok
genObj <- primeCross(genoMVsim, type = "interval", id = "id")
```

```{r map_summary}
#| eval: !expr asreml_ok
n_chr <- length(genObj$geno)
n_int <- sum(sapply(genObj$geno, function(ch) ncol(ch$interval.data)))
n_mk  <- sum(sapply(genObj$geno, function(ch) ncol(ch$data)))
cat(sprintf("Chromosomes : %d\n", n_chr))
cat(sprintf("Markers     : %d\n", n_mk))
cat(sprintf("Intervals   : %d\n", n_int))
cat(sprintf("Lines       : %d\n", nrow(genObj$pheno)))
```

With 594 intervals and 120 lines, the ratio $p/n > 1$ and `qtlAim()` will
automatically use the `vm` (genomic relationship matrix) engine path.

---

## Base Model Development

The starting point for a multivariate analysis is a converged multi-environment
base model that captures the experimental design structure and the genetic
(co)variance across environments. We build this incrementally, comparing three
variance structures for the genetic term.

```{r pheno_setup}
#| eval: !expr asreml_ok
#| results: hide
phenoMVsim$id    <- factor(phenoMVsim$id)
phenoMVsim$Trial <- factor(phenoMVsim$Trial,
                           levels = paste0("Trial", 1:8))
phenoMVsim$Rep   <- factor(phenoMVsim$Rep)
```

### Step 1 — Diagonal structure

The simplest multi-environment model assumes genetic effects in each trial are
**independent**: a diagonal $\boldsymbol{G}_a = \text{diag}(\sigma^2_{g,1},
\ldots, \sigma^2_{g,T})$ with no between-trial genetic correlations. This is
the natural starting point.

```{r base_diag}
#| eval: !expr asreml_ok
#| results: hide
base_diag <- asreml(
  fixed     = yield ~ Trial + Trial:Rep,
  random    = ~ diag(Trial):id,
  residual  = ~ dsum(~units | Trial),
  data      = phenoMVsim,
  na.action = na.method(x = "include"),
  maxit     = 80
)
base_diag <- update(base_diag)
```

### Step 2 — Factor analytic order 1

A single-factor model FA(1) introduces one set of loadings and captures
the dominant axis of genotype-by-environment interaction:

```{r base_fa1}
#| eval: !expr asreml_ok
#| results: hide
base_fa1 <- asreml(
  fixed     = yield ~ Trial + Trial:Rep,
  random    = ~ fa(Trial, 1):id,
  residual  = ~ dsum(~units | Trial),
  data      = phenoMVsim,
  na.action = na.method(x = "include"),
  maxit     = 80
)
base_fa1 <- update(base_fa1)
```

### Step 3 — Factor analytic order 2

A two-factor model FA(2) captures two axes of interaction and provides
a richer description of the genetic covariance structure. This is the model
we will carry into `qtlAim()`:

```{r base_fa2}
#| eval: !expr asreml_ok
#| results: hide
base_fa2 <- asreml(
  fixed     = yield ~ Trial + Trial:Rep,
  random    = ~ fa(Trial, 2):id,
  residual  = ~ dsum(~units | Trial),
  data      = phenoMVsim,
  na.action = na.method(x = "include"),
  maxit     = 80
)
base_fa2 <- update(base_fa2)
```

### Model comparison

We compare the three structures using their restricted log-likelihoods. Each
step adds parameters and the improvement in fit guides the choice of order:

```{r model_comparison}
#| eval: !expr asreml_ok
aic <- function(mod, k) -2 * mod$loglik + 2 * k
# Parameters: diag = 8 var; fa1 = 8 psi + 8 load = 16; fa2 = 8 psi + 16 load = 24
mod_comp <- data.frame(
  Model      = c("diag(Trial)", "fa(Trial, 1)", "fa(Trial, 2)"),
  Parameters = c(8L, 16L, 24L),
  LogLik     = round(c(base_diag$loglik, base_fa1$loglik, base_fa2$loglik), 2),
  AIC        = round(c(aic(base_diag, 8), aic(base_fa1, 16), aic(base_fa2, 24)), 2)
)
print(mod_comp, row.names = FALSE)
```

The FA(2) model is selected as it provides the best balance between fit and
parsimony while capturing two axes of genotype-by-environment interaction
across the eight trials. With 8 trials, FA(2) has 24 free parameters compared
to 36 for the full unstructured matrix.

---

## QTL Mapping — `qtlAim()`

### Running `qtlAim()`

The `Trait` argument activates the multivariate engine, treating `"Trial"` as
the environment factor. Leaving `str = NULL` (the default) instructs
`qtlAim()` to mirror the variance structure of the base model — here FA(2) —
for the composite genome-wide random term, giving consistent modelling of the
genetic covariance at every iteration without needing to respecify the order:

```{r qtlAim}
#| eval: !expr asreml_ok
#| results: hide
qtl_mv <- qtlAim(
  baseModel        = base_fa2,
  genObj           = genObj,
  merge.by         = "id",
  Trait            = "Trial",
  exclusion.window = 20,
  TypeI            = 0.05,
  trace            = FALSE,
  na.action        = na.method(x = "include"),
  maxit            = 80
)
```

```{r qtlAim_class}
#| eval: !expr asreml_ok
#| results: markup
cat("Class        :", paste(class(qtl_mv), collapse = ", "), "\n")
cat("Engine       :", qtl_mv$QTL$path, "\n")
cat("QTL detected :", length(qtl_mv$QTL$qtl), "\n")
cat("QTL keys     :", paste(qtl_mv$QTL$qtl, collapse = ", "), "\n")
```

---

## Results

### `summary()` — final QTL table

The summary table reports one row per environment per detected QTL. The
**Trait** column identifies the environment and a **[MAIN]** or **[INT]**
tag (visible in `print()`) distinguishes QTL whose effects are consistent
across environments from those with significant G×E interaction:

```{r summary_qtl}
#| eval: !expr asreml_ok
summ_mv <- summary(qtl_mv, genObj = genObj)
print(summ_mv)
```

---

## Visualisation — `plot()`

### Genome-wide outlier statistics — `type = "outlier"`

The outlier statistic $o_k = \tilde{\boldsymbol{q}}_k^\top
\boldsymbol{G}_a^{-1}\tilde{\boldsymbol{q}}_k / \tilde{v}_k$ is plotted
across the genome for each forward-selection iteration. Because the statistic
is a quadratic form it is strictly non-negative, and its profile reveals
which chromosomal regions carry the strongest multi-environment signal at
each step:

```{r plot_outlier}
#| eval: !expr asreml_ok
#| fig-height: 10.5
#| fig-cap: "Genome-wide outlier statistics at each forward-selection iteration. The selected QTL position is highlighted; the composite random background is shaded."
plot(qtl_mv, genObj = genObj, type = "outlier", sig.col = "firebrick")
```

### Scaled genome-wide BLUPs — `type = "blups"`

For a multivariate analysis, the BLUPs plot shows one line per trial,
coloured by trial level. This display is particularly informative: intervals
where all trial lines move together indicate main-effect signal, while
intervals where trial lines diverge or cross indicate G×E signal:

```{r plot_blups}
#| eval: !expr asreml_ok
#| fig-height: 10.5
#| fig-cap: "Scaled composite BLUPs across the genome at each iteration, coloured by trial. Convergent lines = consistent main-effect signal; divergent or crossing lines = G×E signal."
plot(qtl_mv, genObj = genObj, type = "blups", sig.col = "firebrick")
```

### Allele effects — `type = "effects"`

The effects lollipop is extended to show one row per trial per QTL. Main-effect
QTL appear as a tight cluster of identically coloured lollipops across all
trials; G×E QTL show variation in magnitude or sign across trials:

```{r plot_effects}
#| eval: !expr asreml_ok
#| fig-height: 7
#| fig-cap: "Per-trial allele effects ± SE for each detected QTL. Firebrick = A allele favoured; steelblue = B allele favoured. Main-effect QTL show consistent colours across trials; G×E QTL show trial-dependent variation."
plot(qtl_mv, genObj = genObj, type = "effects")
```

### Allele contrast — `type = "contrast"`

The contrast plot shows the total genetic value per line against its allele
class at each detected QTL, separately for each trial. Main-effect QTL show
the same direction of separation across all trial panels; crossover G×E QTL
show the separation reversing between panels:

```{r plot_contrast}
#| eval: !expr asreml_ok
#| fig-height: 10
#| fig-cap: "Total genetic value by allele class at each detected QTL, faceted by trial (columns) and QTL (rows). Group means ± SE are overlaid."
plot(qtl_mv, genObj = genObj, type = "contrast", data = phenoData, ncol = 2)
```

---

## Iteration Diagnostics — `aimTrace()`

### LRT trace — `plot = "lrt"`

The mixture chi-squared threshold (shown as a dashed line) reflects the
$T$-dimensional nature of the test — it lies above the univariate $\chi^2_1$
threshold because the test jointly evaluates effects across all $T$
environments:

```{r aimtrace_lrt}
#| eval: !expr asreml_ok
#| fig-height: 4
#| fig-cap: "LRT statistic at each forward-selection iteration. The dashed line is the chi-squared mixture threshold at TypeI = 0.05 for T = 8 environments."
aimTrace(qtl_mv, plot = "lrt", print.out = FALSE)
```

### Effect stability — `plot = "stability"`

The stability plot traces the per-trial allele effect ± 1 SE for each detected
QTL across iterations. Each trial is shown as a separate coloured line. Stable
main-effect QTL will show all trial lines tracking closely together; G×E QTL
will show trial lines that are separated or that diverge as further QTL enter
the model:

```{r aimtrace_stability}
#| eval: !expr asreml_ok
#| fig-height: 6.5
#| fig-cap: "Per-trial allele effect estimates ± 1 SE across iterations, coloured by trial. Stable QTL show consistent per-trial effects; G×E QTL show trial-dependent trajectories."
aimTrace(qtl_mv, plot = "stability", print.out = FALSE, ncol = 2)
```

---

## Summary

The complete multivariate pipeline illustrated in this vignette is:

```r
# 1. Prepare genotype data
genObj <- primeCross(genoMVsim, type = "interval", id = "id")

# 2. Build up the base model
base_diag <- asreml(yield ~ Trial + Trial:Rep,
                    random = ~ diag(Trial):id,
                    residual = ~ dsum(~units | Trial), data = phenoMVsim, ...)
base_fa1  <- asreml(yield ~ Trial + Trial:Rep,
                    random = ~ fa(Trial, 1):id,
                    residual = ~ dsum(~units | Trial), data = phenoMVsim, ...)
base_fa2  <- asreml(yield ~ Trial + Trial:Rep,
                    random = ~ fa(Trial, 2):id,
                    residual = ~ dsum(~units | Trial), data = phenoMVsim, ...)

# 3. Multivariate QTL mapping
qtl_mv <- qtlAim(base_fa2, genObj = genObj, merge.by = "id",
                 Trait = "Trial",
                 na.action = na.method(x = "include"), maxit = 80, ...)

# 4. Results
summary(qtl_mv, genObj = genObj)          # final QTL table (MAIN/INT)
plot(qtl_mv, genObj = genObj, type = "outlier")  # quadratic-form outlier stats
plot(qtl_mv, genObj = genObj, type = "blups")    # multi-trial BLUPs
plot(qtl_mv, genObj = genObj, type = "effects")  # per-trial effect lollipop
plot(qtl_mv, genObj = genObj, type = "contrast", data = phenoData, ncol = 2)
aimTrace(qtl_mv, plot = "lrt",       print.out = FALSE)
aimTrace(qtl_mv, plot = "stability", print.out = FALSE, ncol = 2)
```

---

## References

Verbyla, A.P., Cullis, B.R. & Thompson, R. (2007). The analysis of QTL by
simultaneous use of the full linkage map. *Theoretical and Applied Genetics*,
**116**, 95–111. <https://doi.org/10.1007/s00122-007-0676-4>

Verbyla, A.P., Taylor, J.D. & Verbyla, K.L. (2012). RWGAIM: An efficient
high-dimensional random whole genome average (QTL) interval mapping approach.
*Genetics Research*, **94**, 291–306.
<https://doi.org/10.1017/S0016672312000493>

Taylor, J. & Verbyla, A.P. (2011). R Package wgaim: QTL Analysis in
Bi-Parental Populations Using Linear Mixed Models. *Journal of Statistical
Software*, **40**(7), 1–18. <https://doi.org/10.18637/jss.v040.i07>

Smith, A.B., Cullis, B.R. & Thompson, R. (2001). Analyzing variety by
environment data using multiplicative mixed models and adjustments for spatial
field trend. *Biometrics*, **57**(4), 1138–1147.
<https://doi.org/10.1111/j.0006-341X.2001.01138.x>

Haldane, J.B.S. (1919). The combination of linkage values and the
calculation of distances between the loci of linked factors. *Journal of
Genetics*, **8**(29), 299–309.
