---
title: "Univariate QTL Mapping with wgAim"
author: "Julian Taylor & Ari Verbyla"
date: today
vignette: >
  %\VignetteIndexEntry{Univariate QTL Mapping with wgAim}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    smooth-scroll: true
    embed-resources: true
    theme: litera
    css: custom.css
    highlight-style: github
    code-fold: false
    fig-width: 10
    fig-height: 5
    fig-dpi: 150
    html-math-method: katex
execute:
  warning: false
  message: false
  cache: true
---

```{r setup}
#| include: false
#| cache: false
asreml_ok <- requireNamespace("asreml", quietly = TRUE)
if (requireNamespace("wgAim", quietly = TRUE)) {
  library(wgAim)
} else {
  devtools::load_all("C:/jtaylor/Research/Computing/R Packages/Claude/wgaim",
                     quiet = TRUE)
}
library(ggplot2)
if (asreml_ok) {
  library(asreml)
  asreml.options(trace = FALSE)
}
```

---

## Overview

This vignette demonstrates the complete univariate **wgAim** QTL mapping pipeline for
a biparental mapping population. The pipeline has three stages:

1. **Pre-analysis** — convert a raw `cross` object into the analysis-ready
   `wgCross` format using `primeCross()`.
2. **Analysis** — fit a whole-genome composite model and iteratively select
   QTL using `qtlAim()`.
3. **Post-analysis** — interrogate, summarise, and visualise the detected
   QTL using the `print`, `summary`, and `plot` methods, and refine QTL
   positions with `fineMap()`.

Throughout the vignette we use the `genoRxK` and `phenoRxK` datasets
bundled with the package. These contain marker and phenotypic data from a
254-line Doubled Haploid (DH) wheat population derived from a cross between
varieties **RAC875** and **Kukri**, evaluated at Roseworthy, South Australia
in 2007. The linkage map spans 21 chromosomes with 500 markers. Two traits
are available: grain yield (`yld`, kg/ha) and thousand grain weight (`tgw`).

::: {.note}
**Requires:** ASReml-R V4. The vignette is pre-rendered so no licence is
needed to read it. Run the code chunks only if you have ASReml-R installed.
:::

---

## Mathematical Framework

### The base linear mixed model

Let $\boldsymbol{y}$ be the $n \times 1$ vector of phenotypic observations
for a single trait measured on $n$ lines in a field trial. The starting point
for any `qtlAim()` analysis is a converged **base linear mixed model** (LMM) of the
form

$$
\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\tau}
               + \boldsymbol{Z}_u\boldsymbol{u}
               + \boldsymbol{Z}_g\boldsymbol{g}
               + \boldsymbol{e},
$$

where:

- $\boldsymbol{X}$ is the $n \times p$ design matrix for the **fixed effects**
  $\boldsymbol{\tau}$, which capture known, systematic sources of variation
  such as replicate, range or row trend effects.
- $\boldsymbol{Z}_u$ is the design matrix for **peripheral random effects**
  $\boldsymbol{u}$, representing nuisance sources of variation in the
  experimental layout — for example, incomplete blocks, rows, ranges, or
  other spatial or management groupings. Specifically,
  $\boldsymbol{u} \sim \mathcal{N}(\boldsymbol{0},\, \sigma^2_u\,\boldsymbol{I})$.
- $\boldsymbol{Z}_g$ is the design matrix linking observations to **genetic
  line effects** $\boldsymbol{g}$, with
  $\boldsymbol{g} \sim \mathcal{N}(\boldsymbol{0},\, \sigma^2_g\,\boldsymbol{I})$.
  This term accounts for line-to-line genetic variation prior to any QTL
  modelling.
- $\boldsymbol{e} \sim \mathcal{N}(\boldsymbol{0},\, \sigma^2\,\boldsymbol{R})$
  is the vector of residual errors, where $\boldsymbol{R}$ is a structured
  correlation matrix that captures any remaining spatial or within-trial
  correlation among observations (for example, an AR1$\times$AR1 surface over
  rows and columns).

The base model is fitted and checked for convergence before being passed to
`qtlAim()`. It defines the environmental and design structure that must be
accounted for before the genetic signal can be reliably extracted.

### Introducing additive genetic effects

When QTL mapping is the objective, the genetic term $\boldsymbol{g}$ is
decomposed into two components: an **additive genetic effect** $\boldsymbol{a}$
that captures the cumulative contribution of loci spread across the genome, and
a **residual genetic effect** $\boldsymbol{p}$ that absorbs any remaining
line-level genetic signal not accounted for by the marker/interval data. The
augmented model becomes

$$
\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\tau}
               + \boldsymbol{Z}_u\boldsymbol{u}
               + \boldsymbol{Z}_g\boldsymbol{a}
               + \boldsymbol{Z}_g\boldsymbol{p}
               + \boldsymbol{e},
$$

where $\boldsymbol{a}$ is modelled via the **whole-genome composite term**
described below, and $\boldsymbol{p} \sim \mathcal{N}(\boldsymbol{0},\,
\sigma^2_p\,\boldsymbol{I})$ is the residual polygenic background that
ensures the genetic effects not captured by detected QTL are not forced
into the residual. Retaining $\boldsymbol{p}$ is important: it prevents
inflation of the QTL effects and provides a more honest partition of genetic
variance.

### The whole-genome composite model

`qtlAim()` represents the additive genetic term $\boldsymbol{a}$ as a
**composite genome-wide random effect** that simultaneously models the
contribution of all $p$ marker intervals. Let $\boldsymbol{M}$ be the
$n \times p$ matrix of scaled marker or interval genotype scores
(coded $-1$ / $+1$ for the two parental alleles). The two equivalent
computational paths used are:

- **vm path** ($p > n$): The aggregate marker signal is absorbed into a
  genomic relationship matrix
  $\boldsymbol{K} = \tfrac{1}{p}\boldsymbol{M}\boldsymbol{M}^{\top}$,
  and
  $\boldsymbol{a} \sim \mathcal{N}(\boldsymbol{0},\, \sigma^2_a\,\boldsymbol{K})$.
  This is the standard Gaussian process / GBLUP formulation, efficient when
  there are more markers than lines.
- **mbf path** ($n \geq p$): Each marker effect is treated as an independent
  random regression coefficient,
  $\boldsymbol{a} = \boldsymbol{M}\boldsymbol{q}$ with
  $\boldsymbol{q} \sim \mathcal{N}(\boldsymbol{0},\, \sigma^2_q\,\boldsymbol{I}_p)$,
  equivalent to ridge regression on the marker scores. This path is preferred
  when lines outnumber markers.

Both paths yield identical QTL selection results; the engine selects the
appropriate path automatically based on the dimensions of $\boldsymbol{M}$,
or the user can force the mbf path via `force = TRUE`.

### The likelihood ratio test and forward selection and

Putative QTL are detected by **forward selection**. To confirm the significance
of a putative QTL somewhere on the genome, a **likelihood ratio test** (LRT) is used to compare
the additive LMM against the baseline LMM with both models potetially containing selected QTL
from earlier iterations.

$$
\Lambda = -2\!\left(\ell_{\text{base}} - \ell_{\text{full}}\right)
        \;\stackrel{\text{approx.}}{\sim}\;
        \chi^2_{1},
$$

where $\ell$ denotes the residual log-likelihood. If $\Lambda$ exceeds the
$\chi^2_1$ threshold at the family-wise significance level `TypeI` (default
0.05) a search putative QTL is underatken. No Bonferroni or any other multiple
comparison correction is needed for this test as the `TypeI` value acts like a
family-wise error rate (see Verbyla et al. 2007 for more details.)

Once the null hypothesis of no QTL is rejected a genome-wide **outlier statistic**
for interval $k$ is calculated using

$$
o_k = \frac{\tilde{q}_k^2}{\tilde{v}_k},
$$

for every candidate interval, where $\tilde{q}_k$ is the
posterior mean (BLUP) of the interval effect from the composite random term
and $\tilde{v}_k$ is its posterior variance. The interval with the largest
$o_k$ is the candidate for promotion to a fixed or random QTL.

The confirmed QTL is retained as a fixed or random effect and the composite random term
is rebuilt on the remaining intervals, ensuring subsequent iterations
condition on all previously detected effects. This continues until the null hypothesis is
retained using the LRT threshold, yielding a final model with $s$ fixed or random
QTL effects alongside the remaining additive effects and residual genetic effects.
The final model becomes

$$
\boldsymbol{y} = \boldsymbol{X}\boldsymbol{\tau}
               + \sum_{k=1}^{s} \hat{\beta}_k\,\boldsymbol{x}_k
               + \boldsymbol{Z}_u\boldsymbol{u}
	       + \boldsymbol{Z}_g\boldsymbol{a}_{-s}
               + \boldsymbol{Z}_g\boldsymbol{p}
               + \boldsymbol{e},
$$

where $\hat{\beta}_k$ is the estimated allele substitution effect for the
$k$-th QTL and $\boldsymbol{x}_k$ is the corresponding column of interval
genotype scores and $\boldsymbol{a}_{-s} = \boldsymbol{M}_{-s}\boldsymbol{q}_{-s}$
where the $\boldsymbol{M}_{-s}$ denotes $\boldsymbol{M}$ with an omission of
markers/intervals in the windows of the QTL selected.

---

## Data Preparation — `primeCross()`

`primeCross()` converts a `cross` object from the **qtl** package into a
`wgCross` object ready for `qtlAim()`. Its key task is to construct the
**interval genotype scores** used in interval mapping — Haldane-weighted
midpoint summaries of the flanking marker genotypes that place a virtual
observation at the centre of each inter-marker interval.

```{r load_data}
#| eval: !expr asreml_ok
data(genoRxK,  package = "wgAim")
data(phenoRxK, package = "wgAim")
```

```{r primeCross}
#| eval: !expr asreml_ok
rxk <- primeCross(genoRxK, type = "interval", id = "Genotype")
```

The resulting `wgCross` object stores one list element per chromosome.
Each element contains `$map` (cM positions), `$data` (original marker
genotypes), and `$interval.data` (the derived interval scores). A brief
summary of the map:

```{r map_summary}
#| eval: !expr asreml_ok
n_chr <- length(rxk$geno)
n_int <- sum(sapply(rxk$geno, function(ch) ncol(ch$interval.data)))
n_mk  <- sum(sapply(rxk$geno, function(ch) ncol(ch$data)))
cat(sprintf("Chromosomes : %d\n", n_chr))
cat(sprintf("Markers     : %d\n", n_mk))
cat(sprintf("Intervals   : %d\n", n_int))
```

The full linkage map can be visualised with `linkMap()`:

```{r linkmap_base}
#| eval: !expr asreml_ok
#| fig-height: 10
#| fig-width: 16
#| fig-cap: "RAC875 × Kukri linkage map displayed in three rows. Marker names are shown on each chromosome; tick marks indicate marker positions."
linkMap(rxk, nrow = 3, marker.names = "all") +
  theme(plot.title = element_blank())
```

---

## QTL Mapping — `qtlAim()`

### Base model

The starting point is a converged ASReml-R model that captures the
experimental design without any genomic term. For this trial — a
randomised complete block design with spatial trend — we include a fixed
range effect and random row and genotype terms:

```{r base_model}
#| eval: !expr asreml_ok
#| results: hide
pheno <- phenoRxK
pheno$Genotype <- factor(pheno$Genotype)
pheno$Row      <- factor(pheno$Row)
pheno$Range    <- factor(pheno$Range)

base_yld <- asreml(
  fixed     = yld ~ Range,
  random    = ~ Genotype + Row,
  data      = pheno,
  na.action = na.method(x = "include")
)
base_yld <- update(base_yld)
```

### Running `qtlAim()`

`qtlAim()` takes the fitted base model and the `wgCross` object. The
`merge.by` argument names the column shared between the phenotypic data
and `rxk$pheno` that identifies lines. An exclusion window of 50 cM
prevents closely linked false positives from being selected in consecutive
iterations:

```{r qtlAim}
#| eval: !expr asreml_ok
#| results: hide
qtl_yld <- qtlAim(
  baseModel        = base_yld,
  genObj           = rxk,
  merge.by         = "Genotype",
  exclusion.window = 50,
  TypeI            = 0.05,
  trace            = FALSE
)
```

On completion `qtlAim()` returns an object of class `"qtlAim"` and prints a
brief summary to the console. The engine path (vm or mbf) is chosen
automatically based on the ratio of markers to lines, and the number of
detected QTL reflects how many intervals cleared the LRT threshold:

```{r qtlAim_class}
#| eval: !expr asreml_ok
#| results: markup
cat("Class  :", paste(class(qtl_yld), collapse = ", "), "\n")
cat("Engine :", qtl_yld$QTL$path, "\n")
cat("QTL detected:", length(qtl_yld$QTL$qtl), "\n")
```

---

## Results

### `summary()` — final QTL table

`summary()` collates the final model into a single table with one row per
detected QTL, reporting flanking markers, position, allele effect ± SE,
and cumulative percentage variance explained:

```{r summary_qtl}
#| eval: !expr asreml_ok
summ_yld <- summary(qtl_yld, genObj = rxk)
print(summ_yld)
```

---

## Visualisation — `plot()`

Five complementary plot types are available via the `type` argument of
`plot.qtlAim()`.

### Genome-wide outlier statistics — `type = "outlier"`

The outlier plot shows the genome-wide profile of the interval outlier
statistic $\tilde{q}_k^2 / \tilde{v}_k$ at each iteration, faceted by
iteration number. The selected QTL position is highlighted and the
remaining composite background is shaded. This is the most informative
display for understanding how the forward selection proceeded:

```{r plot_outlier}
#| eval: !expr asreml_ok
#| fig-height: 10.5
#| fig-cap: "Genome-wide outlier statistics at each forward-selection iteration. The vertical red line marks the selected QTL; grey background is the composite random term."
plot(qtl_yld, genObj = rxk, type = "outlier", sig.col = "firebrick")
```

### Scaled genome-wide BLUPs — `type = "blups"`

The BLUPs plot displays the scaled posterior means of all interval effects
from the composite random term at each iteration. Intervals close to a
detected QTL naturally shrink as the QTL absorbs the nearby signal:

```{r plot_blups}
#| eval: !expr asreml_ok
#| fig-height: 10.5
#| fig-cap: "Scaled composite BLUPs across the genome at each iteration."
plot(qtl_yld, genObj = rxk, type = "blups", sig.col = "firebrick")
```

### Allele effects — `type = "effects"`

The effects lollipop summarises the final detected QTL. Each QTL appears
as a lollipop whose head sits at the estimated allele substitution effect
± SE. Colour encodes which allele is favoured: **firebrick** when the A
(RAC875) allele increases yield, **steelblue** when the B (Kukri) allele
does. The right panel shows the percentage of genetic variance explained:

```{r plot_effects}
#| eval: !expr asreml_ok
#| fig-height: 5
#| fig-cap: "Allele effects and percentage variance explained for each detected QTL."
plot(qtl_yld, genObj = rxk, type = "effects")
```

### Allele contrast — `type = "contrast"`

The contrast plot displays the total genetic value for each line against its
allele class at each detected QTL. The total genetic value combines three
components: the composite genome-wide BLUP, the residual line-level genetic
effect, and the summed fixed QTL contributions $\sum_k \hat{\beta}_k x_{ik}$.

This plot requires `phenoData` — the augmented data frame that `qtlAim()`
automatically assigns to your calling environment at the end of the analysis.
`phenoData` contains the original phenotypic columns plus `Gomit` and the
`X.<chr>.<idx>` genotype columns for each detected QTL. The raw `pheno` frame
would not work here as it does not contain those columns.

```{r plot_contrast}
#| eval: !expr asreml_ok
#| fig-height: 8
#| fig-cap: "Total genetic value by allele class at each detected QTL, faceted in two columns. Firebrick = RAC875 (A, +1) allele favoured; steelblue = Kukri (B, −1) allele favoured. Group means ± SE are overlaid."
plot(qtl_yld, genObj = rxk, type = "contrast", data = phenoData, ncol = 2)
```

### Iteration heatmap — `type = "heatmap"`

The heatmap gives a compact genome × iteration view, filling each cell
with the outlier statistic. This makes it easy to see which chromosomal
regions carried persistent signal across multiple iterations:

```{r plot_heatmap}
#| eval: !expr asreml_ok
#| fig-height: 5
#| fig-cap: "Iteration × genome heatmap of the outlier statistic. Darker cells indicate stronger signal."
plot(qtl_yld, genObj = rxk, type = "heatmap", sig.col = "firebrick")
```

### Linkage map overlay — `linkMap()`

`linkMap()` overlays the detected QTL positions onto the genetic map, with
flanking markers bracketing each QTL position:

```{r linkmap_qtl}
#| eval: !expr asreml_ok
#| fig-height: 7.2
#| fig-width: 16
#| fig-cap: "Linkage map with detected QTL positions and marker names. Brackets show flanking markers."
linkMap(qtl_yld, genObj = rxk, marker.names = "all")
```

---

## Iteration Diagnostics — `aimTrace()`

`aimTrace()` traces how the LRT statistic and QTL effect estimates
evolved across iterations, providing a useful diagnostic for assessing
the stability of the detected QTL.

### LRT trace — `plot = "lrt"`

```{r aimtrace_lrt}
#| eval: !expr asreml_ok
#| fig-height: 4
#| fig-cap: "LRT statistic at each forward-selection iteration. The dashed red line is the chi-squared threshold at TypeI = 0.05."
aimTrace(qtl_yld, plot = "lrt", print.out = FALSE)
```

### Effect stability — `plot = "stability"`

```{r aimtrace_stability}
#| eval: !expr asreml_ok
#| fig-height: 6.5
#| fig-cap: "QTL allele effect estimates ± 1 SE across the iterations in which each QTL was present in the model."
aimTrace(qtl_yld, plot = "stability", print.out = FALSE, ncol = 2)
```

---

## Fine Mapping — `fineMap()`

`fineMap()` performs a high-resolution single-QTL scan in a window around
each detected QTL, holding all other detected effects fixed in the model.
For `qtlAim` objects it constructs a dense inferred marker grid across the
window using Haldane interpolation, then fits each candidate interval in
turn and reports the resulting LOD score profile.

We fine-map the four QTL with the largest percentage variance explained,
identified from the summary table. A window of 40 cM and a grid step of
2 cM gives a resolution of approximately 1 cM across each region of
interest:

```{r finemapping}
#| eval: !expr asreml_ok
#| results: hide
fm <- fineMap(qtl_yld, genObj = rxk,
              qtl    = c("Chr.3B.30", "Chr.3D.4", "Chr.3B.10", "Chr.4D.2"),
              window = 40,
              step   = 2,
              exclusion.window = 50)
```

The `plot()` method draws the LOD profile across the window for each
QTL. The filled triangle marks the peak LOD position; the open triangle
shows the original coarser interval position, illustrating how fine
mapping can shift or sharpen the QTL location:

```{r finemapping_plot}
#| eval: !expr asreml_ok
#| fig-height: 8
#| fig-width: 14
#| fig-cap: "Fine-mapping LOD profiles for the four largest QTL arranged in a 2 × 2 grid. Filled triangle: peak fine-map position; open triangle: original qtlAim position."
plot(fm, ncol = 2)
```

---

## Summary

The complete pipeline illustrated in this vignette is:

```r
# 1. Prepare genotype data
rxk      <- primeCross(genoRxK, type = "interval", id = "Genotype")

# 2. Fit base model (no genomic term)
base_yld <- asreml(yld ~ Range, random = ~ Genotype + Row, data = pheno, ...)

# 3. QTL mapping
qtl_yld  <- qtlAim(base_yld, genObj = rxk, merge.by = "Genotype", ...)

# 4. Results
summary(qtl_yld, genObj = rxk)          # final QTL table
plot(qtl_yld,    genObj = rxk, type = "outlier")   # genome profiles
plot(qtl_yld,    genObj = rxk, type = "blups")     # scaled BLUPs
plot(qtl_yld,    genObj = rxk, type = "effects")   # allele effects
plot(qtl_yld,    genObj = rxk, type = "contrast", data = phenoData, ncol = 2)
plot(qtl_yld,    genObj = rxk, type = "heatmap")
aimTrace(qtl_yld, plot = "lrt")
aimTrace(qtl_yld, plot = "stability", ncol = 2)
linkMap(qtl_yld,  genObj = rxk)

# 5. Fine mapping
fm <- fineMap(qtl_yld, genObj = rxk,
              qtl = c("Chr.3B.30", "Chr.3D.4", "Chr.3B.10", "Chr.4D.2"),
              window = 40, step = 2)
plot(fm, ncol = 2)
```

---

## References

Verbyla, A.P., Cullis, B.R. & Thompson, R. (2007). The analysis of QTL by
simultaneous use of the full linkage map. *Theoretical and Applied Genetics*,
**116**, 95–111. <https://doi.org/10.1007/s00122-007-0676-4>

Verbyla, A.P., Taylor, J.D. & Verbyla, K.L. (2012). RWGAIM: An efficient
high-dimensional random whole genome average (QTL) interval mapping approach.
*Genetics Research*, **94**, 291–306.
<https://doi.org/10.1017/S0016672312000493>

Taylor, J. & Verbyla, A.P. (2011). R Package wgaim: QTL Analysis in
Bi-Parental Populations Using Linear Mixed Models. *Journal of Statistical
Software*, **40**(7), 1–18. <https://doi.org/10.18637/jss.v040.i07>

Haldane, J.B.S. (1919). The combination of linkage values and the
calculation of distances between the loci of linked factors. *Journal of
Genetics*, **8**(29), 299–309.
